---
title: "Prediction and Uncertainty analysis"
output: html_document
date: "2024-11-28"
---

```{r}
library(tidyverse)
library(doParallel)
library(ggtext)
library(mgcv)
library(tidymodels)
```

```{r}
load("data/data_R_exported/DS_prepared_for_model_building.RData")
final_mlp_Tricho_nifH_abundance_wf <- 
  readRDS("data/key_model_results/final_mlp_Tricho_nifH_abundance_wf.RDS")
```

- uncertainty 1: MLP model randomness 
- uncertainty 2: light and temperature effects on Q<sub>NCH4</sub> 
- uncertainty 3: MPn to DOP percent 

## Predictiion considering uncertainty caused by mlp randomness

### nifH abundance prediction

```{r nifH abundance prediciton pool}
registerDoParallel()
# this step: about 1 min
mlp_Tricho_nifH_abundance_model_pool <-
  foreach(i = 1:1000) %dopar% {
    set.seed(i) 
    fit(final_mlp_Tricho_nifH_abundance_wf, Tricho_nifH_abundance_data) # Tricho_nifH_abundance_data is from Tricho_nifH_abundance.Rmd
  } 
# nifH prediction, masking land and grids predicted as "no" in rf_Tricho_nifH_existence_prediction.

# landsea_01.msk was downloaded from 
# https://www.ncei.noaa.gov/data/oceans/woa/WOA18/MASKS/landsea_01.msk
mask_land <- 
  read_delim("data/landsea_01.msk", delim = ",", skip = 1) %>%
  arrange(Longitude, Latitude) %>%
  mutate(isLand = Bottom_Standard_Level == 1) %>%
  `[[`("isLand") %>% # True for land, False for sea
  rep(times = 12)

rf_Tricho_nifH_existence_prediction <- 
  readRDS("data/key_model_results/rf_Tricho_nifH_existence_prediction.RDS")

mask_Tricho_existence <- # True for no Tricho, False for Tricho existence
  rf_Tricho_nifH_existence_prediction$.pred_class == "no"

mask <- mask_land | mask_Tricho_existence # True for no Tricho or Land

# this step: about 2 min

mlp_Tricho_nifH_abundance_prediction_pool <-
  foreach(i = 1:1000) %dopar% {
    predict(mlp_Tricho_nifH_abundance_model_pool[[i]], Predictors_for_model_prediction) %>%
      mutate(nifH_abundance_masked = if_else(mask, NA, .pred)) %>%
      dplyr::select(nifH_abundance_masked)
  }

# mlp_Tricho_nifH_abundance_prediction_pool <- 
#   read_rds("data/key_model_results/mlp_Tricho_nifH_abundance_prediction_pool.RDS")

# write_rds(mlp_Tricho_nifH_abundance_prediction_pool, "data/key_model_results/mlp_Tricho_nifH_abundance_prediction_pool.RDS")
```

### nifH abundance prediction: yearly average

```{r}
mask_yearly <- mask %>%
  matrix(ncol = 12, nrow = 180*360) %>%# True for no Tricho or Land
  apply(1, all)

mask_yearly[37380] <- T # existence model predict Yes at month 9, also "sea" in mask_land, but Predictors_for_model_prediction line 180*360*8+37380 contains NA 

# time 1 min
mlp_Tricho_nifH_abundance_prediction_yearly_average_pool <-
  foreach(i = 1:1000) %dopar% {
    nifH_abundance_yearly_average_i <-
    mlp_Tricho_nifH_abundance_prediction_pool[[i]] %>%
      `[[`("nifH_abundance_masked") %>%
      matrix(ncol = 12, nrow = 180*360) %>%
      {`^`(10, .)} %>%
      rowSums(na.rm = T) 
    nifH_abundance_yearly_average_i[mask_yearly] <- NA
    log10(nifH_abundance_yearly_average_i/12)
  }

# mlp_Tricho_nifH_abundance_prediction_yearly_average_pool <-
#   read_rds("data/key_model_results/mlp_Tricho_nifH_abundance_prediction_yearly_average_pool.RDS")

# write_rds(mlp_Tricho_nifH_abundance_prediction_yearly_average_pool, "data/key_model_results/mlp_Tricho_nifH_abundance_prediction_yearly_average_pool.RDS")
```

```{r nifH abundance prediciton summary}
mlp_Tricho_nifH_abundance_prediction_yearly_average_mean <-
  mlp_Tricho_nifH_abundance_prediction_yearly_average_pool %>%
    as.data.frame() %>%
    as.matrix() %>%
    apply(1, mean)
#summary(mlp_Tricho_nifH_abundance_prediction_yearly_average_mean)

mlp_Tricho_nifH_abundance_prediction_yearly_average_median <-
  mlp_Tricho_nifH_abundance_prediction_yearly_average_pool %>%
    as.data.frame() %>%
    as.matrix() %>%
    apply(1, median)

mlp_Tricho_nifH_abundance_prediction_yearly_average_sd <-
  mlp_Tricho_nifH_abundance_prediction_yearly_average_pool %>%
    as.data.frame() %>%
    as.matrix() %>%
    apply(1, sd)
#summary(mlp_Tricho_nifH_abundance_prediction_yearly_average_sd)

mlp_Tricho_nifH_abundance_prediction_yearly_average_CI_upper <-
  mlp_Tricho_nifH_abundance_prediction_yearly_average_pool %>%
    as.data.frame() %>%
    as.matrix() %>%
    apply(1, quantile, probs = 0.975, na.rm = T)

mlp_Tricho_nifH_abundance_prediction_yearly_average_CI_lower <-
  mlp_Tricho_nifH_abundance_prediction_yearly_average_pool %>%
    as.data.frame() %>%
    as.matrix() %>%
    apply(1, quantile, probs = 0.025, na.rm = T)

mlp_Tricho_nifH_abundance_prediction_yearly_average_summary <-
  data.frame(mean = mlp_Tricho_nifH_abundance_prediction_yearly_average_mean,
             median = mlp_Tricho_nifH_abundance_prediction_yearly_average_median,
             sd = mlp_Tricho_nifH_abundance_prediction_yearly_average_sd,
             CI_upper = mlp_Tricho_nifH_abundance_prediction_yearly_average_CI_upper,
             CI_lower = mlp_Tricho_nifH_abundance_prediction_yearly_average_CI_lower)
```

```{r}
## For Matlab plots
# write_csv(mlp_Tricho_nifH_abundance_prediction_yearly_average_summary, "Matlab_plots/mlp_Tricho_nifH_abundance_prediction_yearly_average_summary.csv")

# Tricho_nifH_abundance_data_train %>%
#   mutate(data_type = "train") %>%
#   bind_rows(Tricho_nifH_abundance_data_test %>%
#               mutate(data_type = "test")) %>% # for figure 
#   select(lat_regrid, lon_regrid, Log_Trichodesmium_nifH_integral_regrid, data_type) %>%
#   write_csv("Matlab_plots/Tricho_nifH_abundance_data.csv")

```

### Tricho nfix

```{r Tricho nfix prediction pool}
Calc_Nfix_from_nifH_abundance <- function(nifH_abundance, nifH_per_cell = 10, Nfix_per_cell = 100){
  #nifH in log10(), copies m^-2; nifH_per_cell, nifH copies per cell; Nfix_per_cell, cellular Nfix rate in fmol cell^-2 day^-1
  #return Nfix in umol m^-2 day^-1
  data.frame(nfix = 10^(nifH_abundance$nifH_abundance_masked)/nifH_per_cell*Nfix_per_cell*1e-9)
}

mlp_Tricho_nfix_prediction_pool <- # Nfix in umol m^-2 day^-1
  mlp_Tricho_nifH_abundance_prediction_pool %>%
  map(Calc_Nfix_from_nifH_abundance)

```

```{r}
Days_mat <-
  rep(c(31, 28, 31, 30, 31, 30, 31, 31, 30, 31, 30, 31), each = 360*180) %>%
  matrix(ncol = 12)

# 1 min
mlp_Tricho_nfix_prediction_yearly_pool <-
  foreach(i = 1:1000) %dopar% {
    mlp_Tricho_nfix_prediction_yearly_i <-
      mlp_Tricho_nfix_prediction_pool[[i]] %>%
        `[[`("nfix") %>%
        matrix(ncol = 12) %>%
        `*`(Days_mat) %>%
        rowSums(na.rm = T) # rowSums function return 0 for all NA rows
    mlp_Tricho_nfix_prediction_yearly_i[mask_yearly] <- NA
    mlp_Tricho_nfix_prediction_yearly_i/1000
  }

# mlp_Tricho_nfix_prediction_yearly_pool <-
#   read_rds("data/key_model_results/mlp_Tricho_nfix_prediction_yearly_pool.RDS")
# write_rds(mlp_Tricho_nfix_prediction_yearly_pool, "data/key_model_results/mlp_Tricho_nfix_prediction_yearly_pool.RDS")

```

```{r mlp_Tricho_nfix_prediction_yearly summary}
mlp_Tricho_nfix_prediction_yearly_mean <-
  mlp_Tricho_nfix_prediction_yearly_pool %>%
    as.data.frame() %>%
    as.matrix() %>%
    apply(1, mean)

#mlp_Tricho_nfix_prediction_yearly_mean %>% summary()

mlp_Tricho_nfix_prediction_yearly_median <-
  mlp_Tricho_nfix_prediction_yearly_pool %>%
    as.data.frame() %>%
    as.matrix() %>%
    apply(1, median)

mlp_Tricho_nfix_prediction_yearly_sd <-
  mlp_Tricho_nfix_prediction_yearly_pool %>%
    as.data.frame() %>%
    as.matrix() %>%
    apply(1, sd)

mlp_Tricho_nfix_prediction_yearly_CI_upper <-
  mlp_Tricho_nfix_prediction_yearly_pool %>%
    as.data.frame() %>%
    as.matrix() %>%
    apply(1, quantile, probs = 0.975, na.rm = T)

mlp_Tricho_nfix_prediction_yearly_CI_lower <-
  mlp_Tricho_nfix_prediction_yearly_pool %>%
    as.data.frame() %>%
    as.matrix() %>%
    apply(1, quantile, probs = 0.025, na.rm = T)

mlp_Tricho_nfix_prediction_yearly_summary <-
  data.frame(mean = mlp_Tricho_nfix_prediction_yearly_mean,
             median = mlp_Tricho_nfix_prediction_yearly_median,
             sd = mlp_Tricho_nfix_prediction_yearly_sd,
             CI_upper = mlp_Tricho_nfix_prediction_yearly_CI_upper,
             CI_lower = mlp_Tricho_nfix_prediction_yearly_CI_lower)
```

```{r}
# write_csv(mlp_Tricho_nfix_prediction_yearly_summary, "Matlab_plots/mlp_Tricho_nfix_prediction_yearly_summary.csv")
```

### Total N fixation

```{r Total N fixation}
#mlp_Tricho_nfix_prediction_pool[[1]]
#mlp_Tricho_nfix_prediction_yearly_pool[[1]] # Nfix in mmol m^-2 year^-1
library(raster)
grid_cell_raster <- raster()
area_raster <- area(grid_cell_raster) * 1000*1000
area_matrix <- as.matrix(area_raster)
#area_array <- array(rep(area_matrix[], times = 12), dim = c(180, 360, 12))

# Days_array <-
#   rep(c(31, 28, 31, 30, 31, 30, 31, 31, 30, 31, 30, 31), each = 360*180) %>%
#   array(dim = c(180, 360, 12))

Calc_Total_nfix_yearly <- function(mlp_Tricho_nfix_prediction_yearly){
  mlp_Tricho_nfix_prediction_yearly %>% # mmol m^-2 year^-1 
  matrix(ncol = 360, nrow = 180) %>% 
  `*`(area_matrix) %>% #mmol grid^-2 year^-1
  `*`(1e-3*14*1e-12)  %>% # mmol -> mol -> g -> Tg
  sum(na.rm = T)  # Tg year-1 
}
mlp_Tricho_Total_nifx_pool <- # Tg N year-1
  mlp_Tricho_nfix_prediction_yearly_pool %>%
    map_vec(Calc_Total_nfix_yearly)
summary(mlp_Tricho_Total_nifx_pool)
quantile(mlp_Tricho_Total_nifx_pool, probs = c(0.025, 0.05, 0.1, 0.5, 0.9, 0.95, 0.975))
```

### CH4 production

#### SRP, DOP and SRP/TDP

```{r}
# According to Martiny et al. (2019, Science Advnces), WOA climatoloy overestimates surface DIP more than GLODAP 
# In GLODAP climatoloy dataset, several areas were maked by the authors because of poor data quality. We supplemented it with WOA 2023.

SRP_Glodap_plus_WOA23 <-
  read_csv("data/SRP_Glodap_plus_WOA23.csv", col_names = "SRP") %>%
  `[[`("SRP")

# surface DOP data was derived from Liang et al.(2022, Nature geoscience)

DOP_surface_Liang_model <- 
  read_csv("data/DOP_surface.csv", col_names = "DOP") %>%
  `[[`("DOP") 
  
DOP_to_TDP <- DOP_surface_Liang_model/(SRP_Glodap_plus_WOA23+DOP_surface_Liang_model)

MPn_to_DOP <- 0.08

```

### CH4 production

```{r}
# NCH4_pool.csv contained the ratio of nitrogen fixation to CH4 production under different light and temperature.
# Exp1: current study.
# Exp2: our previous work Zou et al., 2024

NCH4_pool <-
  read_csv("data/NCH4_pool.csv", show_col_types = FALSE)
NCH4_pool %>%
  filter(Experiment == "Exp1",
         Light != 30) %>%
  summarise(mean = mean(NCH4_ratio, na.rm = T))
NCH4_ratio <- 32.39144
```

```{r CH4 production pool}
mlp_Tricho_nfix_prediction_yearly_pool # mmol N m-2 year -1

Cacl_CH4_prod_yearly <- function(mlp_Tricho_nfix_prediction_yearly){
  mlp_Tricho_nfix_prediction_yearly * DOP_to_TDP * MPn_to_DOP *(1/NCH4_ratio)
}

# estimate pool of CH4 production,  mmol CH4 m-2 year-1
Tricho_CH4_prod_yearly_pool <- 
  mlp_Tricho_nfix_prediction_yearly_pool %>%
  map(Cacl_CH4_prod_yearly)
```

```{r CH4 production summary}
# estimate pool of CH4 production: mean, mmol CH4 m-2 year-1
# num [1:64800]
Tricho_CH4_prod_yearly_mean <-
  Tricho_CH4_prod_yearly_pool %>%
    as.data.frame() %>%
    as.matrix() %>%
    apply(1, mean)
```

```{r}
# length(Tricho_CH4_prod_yearly_mean)
# write_csv(data.frame(CH4_prod = Tricho_CH4_prod_yearly_mean), "Matlab_plots/Tricho_CH4_prod_yearly_mean.csv")
```

### Total CH4 prod

```{r}
#mlp_Tricho_nfix_prediction_pool[[1]]
Tricho_CH4_prod_yearly_pool[[1]] # in mmol m^-2 year^-1
grid_cell_raster <- raster()
area_raster <- raster::area(grid_cell_raster) * 1000*1000
area_matrix <- as.matrix(area_raster)

Calc_Total_CH4_prod_yearly <- function(Tricho_CH4_prod_yearly){
  Tricho_CH4_prod_yearly %>% # mmol m^-2 year^-1 
  matrix(ncol = 360, nrow = 180) %>% 
  `*`(area_matrix) %>% #mmol grid^-2 year^-1
  `*`(1e-3*16*1e-12)  %>% # mmol -> mol -> g -> Tg
  sum(na.rm = T)  # Tg year-1 
}

# estimate pool of total yearly CH4 production # Tg CH4 year-1
Tricho_Total_CH4_prod_pool <- 
  Tricho_CH4_prod_yearly_pool %>%
    map_vec(Calc_Total_CH4_prod_yearly)
summary(Tricho_Total_CH4_prod_pool)
quantile(Tricho_Total_CH4_prod_pool, probs = c(0.025, 0.05, 0.1, 0.5, 0.9, 0.95, 0.975))
```

### contribution to CH4 flux

```{r}
# Global_CH4_flux.csv was derived from Weber et al.(2019, Nature communications)
Global_CH4_flux <- 
  read_csv("data/Global_CH4_flux.csv", col_names = "CH4_flux") %>%
  `[[`("CH4_flux") 
Global_CH4_flux[is.na(Tricho_CH4_prod_yearly_pool[[1]])] <- NA

Global_CH4_flux_per_grid <- # mmol CH4 grid-1 year-1
  Global_CH4_flux %>% # mmol CH4-2 year-1
  matrix(nrow = 180, ncol = 360) %>%
  `*`(area_matrix) 
```

```{r}
Tricho_CH4_prod_yearly_pool

Calc_Tricho_CH4_contribution <- function(Tricho_CH4_prod_yearly){
  Tricho_CH4_prod_yearly %>% # mmol m-1 year-1
    matrix(nrow = 180, ncol = 360) %>%
  `*`(area_matrix*100) %>% # mmol grid-1 year-1
  `/`(Global_CH4_flux_per_grid) %>%
  as.vector()# retrun contribution in percent
}

## the ratio of Tricho CH4 production to  CH4 flux : pool
Tricho_CH4_contribution_pool <-
  Tricho_CH4_prod_yearly_pool %>%
  map(Calc_Tricho_CH4_contribution)

## the ratio of Tricho CH4 production to  CH4 flux : mean
Tricho_CH4_contribution_mean <-
  Tricho_CH4_contribution_pool %>%
    as.data.frame() %>%
    as.matrix() %>%
    apply(1, mean)

summary(Tricho_CH4_contribution_mean)
quantile(Tricho_CH4_contribution_mean, probs = c(0.9025, 0.95, 0.9975), na.rm = T)

```

```{r}
# 束毛藻甲烷产量占当地甲烷通量的超过10%的网格点比例 集合
mask_yearly #True for no Tricho or Land 
Tricho_CH4_contribution_pool
.x <- Tricho_CH4_contribution_pool[[1]]
Calc_Tricho_CH4_contribution_10 <- function(.x) {
  Tricho_yes <- .x[!mask_yearly]
  mean(Tricho_yes >= 10, na.rm = T)
}

Tricho_CH4_contribution_10_pool <-
  map_dbl(Tricho_CH4_contribution_pool, Calc_Tricho_CH4_contribution_10)

summary(Tricho_CH4_contribution_10_pool)
quantile(Tricho_CH4_contribution_10_pool, probs = c(0.025, 0.05, 0.1, 0.5, 0.9, 0.95, 0.975))
```

```{r}
# write_csv(data.frame(Tricho_contri = Tricho_CH4_contribution_mean), "Matlab_plots/Tricho_CH4_contri_yearly_mean.csv")
```

#### total contribution

```{r}
Tricho_Total_CH4_prod_pool # Tg year-1

Global_CH4_flux_per_grid[mask_yearly] <- NA # mmol CH4 grid-1 year-1
Tricho_total_CH4_contribution_pool <- 
  Tricho_Total_CH4_prod_pool/sum(Global_CH4_flux_per_grid*1e-3*16*1e-12, na.rm = T)*100

summary(Tricho_total_CH4_contribution_pool)
quantile(Tricho_total_CH4_contribution_pool, probs = c(0.025, 0.05, 0.1, 0.5, 0.9, 0.95, 0.975))
```

# Uncertainty and senstivity analysis

1.  Light and temperature effect
2.  MPn to DOP ratio [0.06, 0.1]
3.  SRP uncertainty (optional) \## Light

```{r}
#fig_PARmld_histogram <- 
Predictors_for_model_prediction %>%
  mutate(kd = 0.0665 + 0.874*kd_490 - 0.00121/kd_490,
         PAR_mld = PAR/(kd*MLD) * (1 - exp(-(kd*MLD))),
         Tricho_nifH_existence = rf_Tricho_nifH_existence_prediction$.pred_class) %>%
  filter(rf_Tricho_nifH_existence_prediction$.pred_class == "yes") %>%
  select(lat_regrid, lon_regrid, Sampling_month, PAR_mld, MLD, PAR, kd_490) %>%
  drop_na() %>%
  #write_rds("data/PARmld_histogram.RDS")
  #filter(PAR_mld < 5.184)
  #filter(PAR_mld > 31.968)
  #summary()
  ggplot() +
  geom_histogram(aes(x = PAR_mld), fill = "#69b3a2",color="#e9ecef", alpha=0.9, bins = 30) +
  annotate("segment", x = 5.184, xend = 5.184, y = 10000, yend = 5000, arrow = arrow(length = unit(0.2, "cm")), linewidth = 0.5, linetype = "dashed") +
  geom_richtext(aes(x = x, y = y, label = text), data = data.frame(x = 5.184, y = 11000, text = "5.2 Einstein m<sup>-2</sup> day<sup>-1</sup>, corresponding to<br>60 &mu;mol photons m<sup>-2</sup> s<sup>-1</sup> with 12:12 h light/dark cycle"), size =2.5, family = "serif")+
  annotate("segment", x = 31.968, xend = 31.968, y = 10000, yend = 5000, arrow = arrow(length = unit(0.2, "cm")), linewidth = 0.5, linetype = "dashed") +
  geom_richtext(aes(x = x, y = y, label = text), data = data.frame(x = 31.968, y = 11000, text = "32.0 Einstein m<sup>-2</sup> day<sup>-1</sup>, corresponding to<br>370 &mu;mol photons m<sup>-2</sup> s<sup>-1</sup> with 12:12 h light/dark cycle"), size =2.5, family = "serif")+
  scale_x_continuous(breaks = c(0, 10, 20, 30, 40, 50)) +
  xlab("The average PAR in the mixed layer (Einstein m<sup>-2</sup> day<sup>-1</sup>)") +
  ylab("Number of data points") +
  theme_light() +
  theme(axis.title.x = element_markdown(family = "serif", size = 16),
        axis.title.y = element_markdown(family = "serif", size = 16),
        axis.text = element_text(family = "serif", size = 14)
        )
```

```{r}
ggsave("figures/histogram_PAR_mld.jpg",dpi = 600, width = 8, height = 5)
```

```{r}
885/132684
8245/132684
```

> The light setup in our controlled experiments covers most of light intensities experienced by *Trichodesmium*.

## Temperature

```{r}
fig_sst_histgram <-
Predictors_for_model_prediction %>%
  mutate(Tricho_nifH_existence = rf_Tricho_nifH_existence_prediction$.pred_class) %>%
  filter(rf_Tricho_nifH_existence_prediction$.pred_class == "yes") %>%
  select(lat_regrid, lon_regrid, Sampling_month, SST) %>%
  drop_na() %>%
  #filter(PAR_mld < 5.184)
  #filter(PAR_mld > 31.968)
  #summary() %>%
  ggplot() +
  geom_histogram(aes(x = SST), fill = "#69b3a2",color="#e9ecef", alpha=0.9, bins = 40) +
  scale_x_continuous(breaks = c(16, 20, 23, 27, 31, 35)) +
  xlab("SST (&deg;C)") +
  ylab("Number of data points")+
  theme_light() +
  theme(axis.title.x = element_markdown(family = "serif", size = 16),
        axis.title.y = element_markdown(family = "serif", size = 16),
        axis.text = element_text(family = "serif", size = 14))
```

```{r}
ggsave("figures/histogram_SST.jpg",dpi = 600, width = 8, height = 5)

```

```{r}
Predictors_for_model_prediction %>%
  mutate(Tricho_nifH_existence = rf_Tricho_nifH_existence_prediction$.pred_class) %>%
  filter(rf_Tricho_nifH_existence_prediction$.pred_class == "yes") %>%
  select(lat_regrid, lon_regrid, Sampling_month, SST) %>%
  drop_na() %>%
  write_rds("data/sst_histogram.RDS")

fig_PARmld_histogram/fig_sst_histgram

Predictors_for_model_prediction %>%
  mutate(Tricho_nifH_existence = rf_Tricho_nifH_existence_prediction$.pred_class) %>%
  filter(rf_Tricho_nifH_existence_prediction$.pred_class == "yes") %>%
  select(lat_regrid, lon_regrid, Sampling_month, SST) %>%
  drop_na() %>%
  filter(SST >= 20,
         SST<= 27)

58176/132697
```

### build realationship between temperature and NCH4

```{r}
NCH4_pool <-
  read_csv("data/NCH4_pool.csv", show_col_types = FALSE)
NCH4_pool %>%
  filter(Light != 30) %>%
  drop_na() %>%
  #mutate(NCH4_ratio_pred = predict(NCH4_ratio_temperature_gam)) %>%
  ggplot() +
  geom_point(aes(Temperature, NCH4_ratio, color = Data_source))
```

```{r using GAM}
NCH4_ratio_temperature_gam <-
  gam(NCH4_ratio ~ s(Temperature, k = 5), data = NCH4_pool %>% filter(Light != 30) %>% drop_na())
# NCH4_pool %>%
#   write_rds("data/NCH4_pool.RDS")
predict(NCH4_ratio_temperature_gam, se.fit = T, newdata = list(Temperature = seq(from = 16, to = 31, by = 0.1))) %>%
  #write_rds("data/NCH4_temperature_gam.RDS")
  as.data.frame() %>%
  mutate(Temperature = seq(from = 16, to = 31, by = 0.1)) %>%
  ggplot() + 
  geom_line(aes(Temperature, fit), linetype= 2) +
  geom_ribbon(aes(x = Temperature, y = fit, ymin = fit - 1.96*se.fit, ymax = fit + 1.96*se.fit), alpha = 0.5) +
  geom_point(aes(Temperature, NCH4_ratio, shape = Data_source), size = 3, data = NCH4_pool %>% filter(Light != 30) %>% drop_na(), position = position_jitter(width = 0.1)) +
  scale_shape_manual(values = c(1, 2))+
  ylab("Q<sub>N_CH<sub>4</sub></subsub>") +
  xlab("Tempearture (&deg;C)") +
  theme_light() +
  theme(axis.title.x = element_markdown(size = 16, family = "serif"),
        axis.title.y = element_markdown(size = 16, family = "serif"),
        axis.text = element_text(size = 14, family = "serif"),
        legend.text = element_text(size = 12, family = "serif"),
        legend.position = "inside", 
        legend.position.inside = c(0.7, 0.9),
        legend.title = element_blank(),
        legend.background = element_rect(color = "black", linewidth = 0.2))
ggsave("figures/QNCH4_temperature.png",dpi = 600, width = 8, height = 5)
```

### Simulating 1000 sets of NCH4_ratio

reference:<https://mikl.dk/post/2019-prediction-intervals-for-gam/>

```{r}
gam_beta <- coef(NCH4_ratio_temperature_gam)
gam_V <- vcov(NCH4_ratio_temperature_gam)
num_beta_vecs <- 1000
gam_Cv <- chol(gam_V)
set.seed(1)
gam_nus <- rnorm(num_beta_vecs*length(gam_beta))
gam_beta_sims <-
  gam_beta + t(gam_Cv) %*% matrix(gam_nus, nrow = length(gam_beta), ncol = num_beta_vecs)
d_beta <- cbind(summary(NCH4_ratio_temperature_gam)$se, apply(gam_beta_sims, 1, sd))
Predictors_SST <- Predictors_for_model_prediction %>%
  select(SST) %>%
  rename(Temperature = SST)
covar_sim <- predict(NCH4_ratio_temperature_gam, newdata = Predictors_SST, type = "lpmatrix")
linpred_sim <- covar_sim %*% gam_beta_sims
invlink <- function(x) x
exp_val_sim <- invlink(linpred_sim)
NCH4_ratio_sim <-
   matrix(rnorm(n = prod(dim(exp_val_sim)), 
                mean = exp_val_sim, 
                sd = sqrt(summary(NCH4_ratio_temperature_gam)$scale)), 
          nrow = nrow(exp_val_sim), 
          ncol = ncol(exp_val_sim))

NCH4_ratio_sim[NCH4_ratio_sim <=0 ] <- NA
dim(NCH4_ratio_sim)
Predictors_SST %>% dim()

NCH4_ratio_sim[9:23, 3:5]
```

### Simulating 1000 sets of MPn_to_DOP ratio

```{r}
set.seed(4)
MPn_to_DOP_sim <-
  runif(prod(dim(NCH4_ratio_sim)), min = 0.06, max = 0.1) %>%
  matrix(ncol = 1000)

DOP_to_TDP_sim <-
  rep(DOP_to_TDP, times = 12 * 1000) %>%
  matrix(nrow = 180*360*12, ncol = 1000)

Days_sim <- 
  rep(c(31, 28, 31, 30, 31, 30, 31, 31, 30, 31, 30, 31), each = 360*180) %>%
  rep(times = 1000) %>%
  matrix(nrow = 180*360*12, ncol = 1000)

dim(MPn_to_DOP_sim)
dim(DOP_to_TDP_sim)
dim(Days_sim)
```

```{r}
length(mlp_Tricho_nifH_abundance_prediction_pool)
mlp_Tricho_nifH_abundance_prediction_pool[[1]] %>% str()

Tricho_CH4_prod_yearly_sim_mat <- 
  mlp_Tricho_nifH_abundance_prediction_pool %>%
    as.data.frame() %>%
    as.matrix() %>%
    {`^`(10, .)} %>% # copies m-2
    `*`(100/10) %>% # nfix fmol N m-2 day-1
    `/`(NCH4_ratio_sim) %>%
    `*`(MPn_to_DOP_sim) %>%
    `*`(DOP_to_TDP_sim) %>% # CH4 fmol m-2 day-1
    `*`(Days_sim) %>% #CH4 fmol m-2 month-1
    array(dim = c(180*360, 12, 1000)) %>%
    apply(c(1,3), sum, na.rm = T) %>% # CH4 fmol m-2 year-1
    `*`(1e-12) # CH4 mmol m-2 year-1
```

```{r}
dim(Tricho_CH4_prod_yearly_sim_mat)

Tricho_CH4_prod_yearly_sim_mean <- 
  Tricho_CH4_prod_yearly_sim_mat %>%
    rowMeans(na.rm = T)
Tricho_CH4_prod_yearly_sim_mean[mask_yearly] <- NA
```

```{r}
# write_csv(data.frame(CH4_prod = Tricho_CH4_prod_yearly_sim_mean),
#           "Matlab_plots/Tricho_CH4_prod_yearly_sim_mean.csv")

```

```{r}
Tricho_CH4_contribution_sim_mat <-
  Tricho_CH4_prod_yearly_sim_mat %>%
    apply(2, Calc_Tricho_CH4_contribution)

Tricho_CH4_contribution_sim_mean <-
  Tricho_CH4_contribution_sim_mat %>%
  apply(1, mean, na.rm = T)
Tricho_CH4_contribution_sim_mean[mask_yearly] <- NA
Tricho_CH4_contribution_sim_mean %>% summary()

```

```{r}
# write_csv(data.frame(CH4_contri = Tricho_CH4_contribution_sim_mean),
#           "Matlab_plots/Tricho_CH4_contribution_sim_mean.csv")

```

#### Total CH4 prod and contribution simulation

```{r}
Tricho_Total_CH4_prod_sim_pool <-
  Tricho_CH4_prod_yearly_sim_mat %>%
    as.data.frame() %>%
    map_vec(Calc_Total_CH4_prod_yearly)


summary(Tricho_Total_CH4_prod_sim_pool)
quantile(Tricho_Total_CH4_prod_sim_pool, probs = c(0.025, 0.05, 0.1, 0.5, 0.9, 0.95, 0.975))

```

```{r}
Global_CH4_flux_per_grid[mask_yearly] <- NA # mmol CH4 grid-1 year-1
Tricho_total_CH4_contribution_sim_pool <- 
  Tricho_Total_CH4_prod_sim_pool/sum(Global_CH4_flux_per_grid*1e-3*16*1e-12, na.rm = T)*100

summary(Tricho_total_CH4_contribution_sim_pool)
quantile(Tricho_total_CH4_contribution_sim_pool, probs = c(0.025, 0.05, 0.1, 0.5, 0.9, 0.95, 0.975))
```
