---
title: "Tricho_nifH_abundance"
output: html_document
date: "2024-11-12"
---

> -   In this Rmarkdown file, random forest (RF) and multilayer perceptron (MLP) models were built and optimized to predict the global *Trichodesmium* *nifH* gene abundance using `tidymodels` framework.
> -   It took \~ 30 min to execute this script on our server.

```{r}
library(tidyverse)
library(tidymodels)
library(raster)
library(doParallel)
registerDoParallel()
```

```{r}
# See `Dataset_preparation.R` for information of `DS_prepared_for_model_building.RData`

load("data/data_R_exported/DS_prepared_for_model_building.RData") 

# See `Tricho_nifH_existence.Rmd` for information of `rf_Tricho_nifH_existence_prediction.RDS`
rf_Tricho_nifH_existence_prediction <-
  read_rds("data/key_model_results/rf_Tricho_nifH_existence_prediction.RDS")
```

## Training, testing and validation dataset

```{r}
Tricho_nifH_abundance_data <-
  DS_for_ML_building_Tricho_nifH %>%
    filter(Trichodesmium_nifH_integral_regrid > 0) %>%
    mutate(Log_Trichodesmium_nifH_integral_regrid = log10(Trichodesmium_nifH_integral_regrid*1e6)) %>%
    dplyr::select(-Trichodesmium_nifH_integral_regrid)

set.seed(27)
Tricho_nifH_abundance_data_split <- 
  Tricho_nifH_abundance_data %>% 
  initial_split(strata = "Log_Trichodesmium_nifH_integral_regrid")

Tricho_nifH_abundance_data_train <-
  training(Tricho_nifH_abundance_data_split)

Tricho_nifH_abundance_data_test <-
  testing(Tricho_nifH_abundance_data_split)

set.seed(94)
Tricho_nifH_abundance_data_folds <- vfold_cv(Tricho_nifH_abundance_data_train, v = 10, repeats = 5)

```

## RF models

### Set up the `workflow` for model building, hyperparameter tuning and prediction

```{r}
rf_Tricho_nifH_abundance_rec <-
  recipe(Log_Trichodesmium_nifH_integral_regrid ~ ., data = Tricho_nifH_abundance_data_train) %>%
  step_log(Chla, DIN, DIP, kd_490, MLD, offset = 0.1 ) %>%
  step_normalize(all_numeric_predictors()) %>%
  update_role(lat_regrid, lon_regrid, Sampling_month, new_role = "ID") # original lat, lon and month were not used for the training

rf_Tricho_nifH_abundance_spec <- rand_forest(
  mtry = tune(),
  trees = tune(),
  min_n = tune()) %>%
  set_mode("regression") %>%
  set_engine("ranger")

rf_Tricho_nifH_abundance_wf <- 
  workflow() %>%
  add_recipe(rf_Tricho_nifH_abundance_rec) %>%
  add_model(rf_Tricho_nifH_abundance_spec)
```

### Tuning hyperparameters

#### Coarse tuning

```{r coarse tune}
set.seed(453)
rf_Tricho_nifH_abundance_tune <- tune_grid(
  rf_Tricho_nifH_abundance_wf,
  resamples = Tricho_nifH_abundance_data_folds,
  grid = 100
)
autoplot(rf_Tricho_nifH_abundance_tune)
```

#### Fine tuning

```{r fine tune}
set.seed(124)
rf_Tricho_nifH_abundance_tune_grid <- 
  grid_space_filling(
    mtry(range = c(1, 14)),
    trees(range = c(1, 2000)),
    min_n(range = c(2, 10)),
    size = 100
  )

set.seed(14)
rf_Tricho_nifH_abundance_tune_fine <- tune_grid(
  rf_Tricho_nifH_abundance_wf,
  resamples = Tricho_nifH_abundance_data_folds,
  grid = rf_Tricho_nifH_abundance_tune_grid,
  control = control_grid(parallel_over = "everything")
)
autoplot(rf_Tricho_nifH_abundance_tune_fine)
```

```{r}
set.seed(24)
rf_Tricho_nifH_abundance_tune_grid1 <- 
  grid_space_filling(
    mtry(range = c(1, 14)),
    trees(range = c(10, 2000)),
    min_n(range = c(2, 10)),
    size = 100
  )

set.seed(314)
rf_Tricho_nifH_abundance_tune_fine1 <- tune_grid(
  rf_Tricho_nifH_abundance_wf,
  resamples = Tricho_nifH_abundance_data_folds,
  grid = rf_Tricho_nifH_abundance_tune_grid1
)
autoplot(rf_Tricho_nifH_abundance_tune_fine1)
```

```{r}
set.seed(124)
rf_Tricho_nifH_abundance_tune_grid2 <- 
  grid_space_filling(
    mtry(range = c(1, 14)),
    trees(range = c(50, 2000)),
    min_n(range = c(2, 10)),
    size = 100
  )

set.seed(34)
rf_Tricho_nifH_abundance_tune_fine2 <- tune_grid(
  rf_Tricho_nifH_abundance_wf,
  resamples = Tricho_nifH_abundance_data_folds,
  grid = rf_Tricho_nifH_abundance_tune_grid2
)
autoplot(rf_Tricho_nifH_abundance_tune_fine2)
```

```{r}
set.seed(14)
rf_Tricho_nifH_abundance_tune_grid3 <- 
  grid_space_filling(
    mtry(range = c(1, 8)),
    trees(range = c(50, 2000)),
    min_n(range = c(2, 10)),
    size = 100
  )

set.seed(334)
rf_Tricho_nifH_abundance_tune_fine3 <- tune_grid(
  rf_Tricho_nifH_abundance_wf,
  resamples = Tricho_nifH_abundance_data_folds,
  grid = rf_Tricho_nifH_abundance_tune_grid3
)
autoplot(rf_Tricho_nifH_abundance_tune_fine3)
```

```{r}
set.seed(445)
rf_Tricho_nifH_abundance_tune_grid4 <- 
  grid_space_filling(
    mtry(range = c(1, 8)),
    trees(range = c(50, 2000)),
    min_n(range = c(2, 6)),
    size = 100
  )

set.seed(343)
rf_Tricho_nifH_abundance_tune_fine4 <- tune_grid(
  rf_Tricho_nifH_abundance_wf,
  resamples = Tricho_nifH_abundance_data_folds,
  grid = rf_Tricho_nifH_abundance_tune_grid4
)
autoplot(rf_Tricho_nifH_abundance_tune_fine4)
```

```{r}
set.seed(453)
rf_Tricho_nifH_abundance_tune_grid5 <- 
  grid_space_filling(
    mtry(range = c(2, 8)),
    trees(range = c(100, 2000)),
    min_n(range = c(2, 6)),
    size = 100
  )

set.seed(452)
rf_Tricho_nifH_abundance_tune_fine5 <- tune_grid(
  rf_Tricho_nifH_abundance_wf,
  resamples = Tricho_nifH_abundance_data_folds,
  grid = rf_Tricho_nifH_abundance_tune_grid5
)
autoplot(rf_Tricho_nifH_abundance_tune_fine5)
```

### Selecting the best RF model

```{r}
final_rf_Tricho_nifH_abundance_wf <- 
  rf_Tricho_nifH_abundance_wf %>%
  finalize_workflow(select_best(rf_Tricho_nifH_abundance_tune_fine5, metric = "rmse"))

set.seed(41)
rf_Tricho_nifH_abundance_last_fit <-
  last_fit(final_rf_Tricho_nifH_abundance_wf, Tricho_nifH_abundance_data_split)

# performance on testing dataset 
collect_metrics(rf_Tricho_nifH_abundance_last_fit)

```

```{r}
# figures of predicted vs. observed
fig_rf_test_performance <-
collect_predictions(rf_Tricho_nifH_abundance_last_fit) %>%
  ggplot(aes(Log_Trichodesmium_nifH_integral_regrid, .pred)) +
  geom_abline(lty = 2, color = "gray50") +
  geom_point(alpha = 0.5, color = "midnightblue") +
  coord_fixed() +
  ylab("Predicted *nifH* abundance (log10 copies m<sup>-2</sup>)") +
  xlab("Observed *nifH* abundance (log10 copies m<sup>-2</sup>)") +
  #annotate("text", x = 6.5, y = 12, label = "b") +
  annotate("text", x = 8, y = 11.5, label = "Random Forest model", family = "serif", size = 6) +
  scale_x_continuous(limits = c(6, 12)) +
  scale_y_continuous(limits = c(6, 12)) +
  theme_light() +
  theme(axis.title.x = element_markdown(size = 14, family = "serif"),
        #axis.title.y = element_markdown(size = 16, family = "serif"),
        axis.title.y = element_blank(),
        axis.text = element_text(size = 14, family = "serif"),
        )
#ggsave("figures/RF_test_performance.jpg",dpi = 600, width = 8, height = 5)
```

### write key results to disk

```{r}
# write_rds(final_rf_Tricho_nifH_abundance_wf, file = "data/key_model_results/final_rf_Tricho_nifH_abundance_wf.RDS")
# 
# write_rds(rf_Tricho_nifH_abundance_last_fit, file = "data/key_model_results/rf_Tricho_nifH_abundance_last_fit.RDS")
```

## mlp models

### Set up the `workflow` for model building, hyperparameter tuning and prediction

```{r}
mlp_Tricho_nifH_abundance_spec <- 
# Three hyperparameters were tuned: 
# 1. epochs: Epochs 
# 2. hidden_units: number of hidden units 
# 3. penalty: Amount of Regularization  
  mlp(epochs = tune(), 
      hidden_units = tune(),  
      penalty = tune()
      ) %>%
  set_mode("regression") %>% 
  set_engine("nnet") 

mlp_Tricho_nifH_abundance_rec <-
  recipe(Log_Trichodesmium_nifH_integral_regrid ~ ., data = Tricho_nifH_abundance_data_train) %>%
  step_log(Chla, DIN, DIP, kd_490, MLD, offset = 0.1 ) %>%
  step_normalize(all_numeric_predictors()) %>%
  update_role(lat_regrid, lon_regrid, Sampling_month, new_role = "ID")  

mlp_Tricho_nifH_abundance_wf <- 
  workflow() %>%
  add_recipe(mlp_Tricho_nifH_abundance_rec) %>%
  add_model(mlp_Tricho_nifH_abundance_spec)
```

### Tuning hyperparameters

#### Coarse tuning

```{r}
set.seed(342)
mlp_Tricho_nifH_abundance_tuned <- tune_grid(
  mlp_Tricho_nifH_abundance_wf,
  resamples = Tricho_nifH_abundance_data_folds,
  grid = 100
)
autoplot(mlp_Tricho_nifH_abundance_tuned) 
```

#### Fine tuning

```{r}
set.seed(431)
mlp_Tricho_nifH_abundance_grid_fine <- 
  grid_space_filling(
    epochs(range = c(10, 1000)),
    hidden_units(range = c(1, 10)), 
    penalty(range = c(-2.5, 0)),
    size = 100
  )

set.seed(789)
mlp_Tricho_nifH_abundance_tuned_fine <- tune_grid(
  mlp_Tricho_nifH_abundance_wf,
  resamples = Tricho_nifH_abundance_data_folds,
  grid = mlp_Tricho_nifH_abundance_grid_fine
)
autoplot(mlp_Tricho_nifH_abundance_tuned_fine)
```

```{r}
set.seed(177)
mlp_Tricho_nifH_abundance_grid_fine1 <- 
  grid_space_filling(
    epochs(range = c(10, 1000)),
    hidden_units(range = c(3, 15)), 
    penalty(range = c(-2.5, 0)),
    size = 100
  )

set.seed(89)
mlp_Tricho_nifH_abundance_tuned_fine1 <- tune_grid(
  mlp_Tricho_nifH_abundance_wf,
  resamples = Tricho_nifH_abundance_data_folds,
  grid = mlp_Tricho_nifH_abundance_grid_fine1
)
autoplot(mlp_Tricho_nifH_abundance_tuned_fine1)
```

```{r}
set.seed(177)
mlp_Tricho_nifH_abundance_grid_fine2 <- 
  grid_space_filling(
    epochs(range = c(10, 1000)), #epochs(range = c(300, 500)),
    hidden_units(range = c(3, 15)), # set tune parameters range based on last tune result.
    penalty(range = c(-1.5, 0)),
    size = 100
  )

set.seed(89)
mlp_Tricho_nifH_abundance_tuned_fine2 <- tune_grid(
  mlp_Tricho_nifH_abundance_wf,
  resamples = Tricho_nifH_abundance_data_folds,
  grid = mlp_Tricho_nifH_abundance_grid_fine2
)
autoplot(mlp_Tricho_nifH_abundance_tuned_fine2)
```

```{r}
set.seed(79)
mlp_Tricho_nifH_abundance_grid_fine3 <- 
  grid_space_filling(
    epochs(range = c(100, 1000)),
    hidden_units(range = c(3, 15)), # set tune parameters range based on last tune result.
    penalty(range = c(-1.5, 0)),
    size = 100
  )

set.seed(9)
mlp_Tricho_nifH_abundance_tuned_fine3 <- tune_grid(
  mlp_Tricho_nifH_abundance_wf,
  resamples = Tricho_nifH_abundance_data_folds,
  grid = mlp_Tricho_nifH_abundance_grid_fine3
)

autoplot(mlp_Tricho_nifH_abundance_tuned_fine3)
```

```{r}
set.seed(791)
mlp_Tricho_nifH_abundance_grid_fine4 <- 
  grid_space_filling(
    epochs(range = c(100, 1000)),
    hidden_units(range = c(3, 15)), # set tune parameters range based on last tune result.
    penalty(range = c(-0.8, 0)),
    size = 100
  )

set.seed(93)
mlp_Tricho_nifH_abundance_tuned_fine4 <- tune_grid(
  mlp_Tricho_nifH_abundance_wf,
  resamples = Tricho_nifH_abundance_data_folds,
  grid = mlp_Tricho_nifH_abundance_grid_fine4
)

autoplot(mlp_Tricho_nifH_abundance_tuned_fine4)
```

```{r}
set.seed(71)
mlp_Tricho_nifH_abundance_grid_fine5 <- 
  grid_space_filling(
    epochs(range = c(100, 1000)),
    hidden_units(range = c(9, 20)), # set tune parameters range based on last tune result.
    penalty(range = c(-0.8, 0)),
    size = 100
  )

set.seed(33)
mlp_Tricho_nifH_abundance_tuned_fine5 <- tune_grid(
  mlp_Tricho_nifH_abundance_wf,
  resamples = Tricho_nifH_abundance_data_folds,
  grid = mlp_Tricho_nifH_abundance_grid_fine5
)

autoplot(mlp_Tricho_nifH_abundance_tuned_fine5)
```

```{r}
set.seed(91)
mlp_Tricho_nifH_abundance_grid_fine5 <- 
  grid_space_filling(
    epochs(range = c(100, 1000)),
    hidden_units(range = c(9, 20)), # set tune parameters range based on last tune result.
    penalty(range = c(-0.5, 0)),
    size = 100
  )

set.seed(923)
mlp_Tricho_nifH_abundance_tuned_fine5 <- tune_grid(
  mlp_Tricho_nifH_abundance_wf,
  resamples = Tricho_nifH_abundance_data_folds,
  grid = mlp_Tricho_nifH_abundance_grid_fine5
)

autoplot(mlp_Tricho_nifH_abundance_tuned_fine5)
```

### Selecting the best MLP model

```{r}
final_mlp_Tricho_nifH_abundance_wf <- 
  mlp_Tricho_nifH_abundance_wf %>%
  finalize_workflow(select_best(mlp_Tricho_nifH_abundance_tuned_fine5, metric = "rmse"))

set.seed(76)
mlp_Tricho_nifH_abundance_last_fit <-
  last_fit(final_mlp_Tricho_nifH_abundance_wf, Tricho_nifH_abundance_data_split)
collect_metrics(mlp_Tricho_nifH_abundance_last_fit)

# performance on testing dataset
collect_metrics(rf_Tricho_nifH_abundance_last_fit)
```

```{r}
# figure of predicted vs. observed
fig_mlp_test_performance <-
collect_predictions(mlp_Tricho_nifH_abundance_last_fit) %>%
  ggplot(aes(Log_Trichodesmium_nifH_integral_regrid, .pred)) +
  geom_abline(lty = 2, color = "gray50") +
  geom_point(alpha = 0.5, color = "midnightblue") +
  coord_fixed() +
  ylab("Predicted *nifH* abundance (log10 copies m<sup>-2</sup>)") +
  xlab("Observed *nifH* abundance (log10 copies m<sup>-2</sup>)") +
  #annotate("text", x = 6.5, y = 12, label = "a") +
  annotate("text", x = 7, y = 11.5, label = "MLP model", family = "serif", size = 6) +
  scale_x_continuous(limits = c(6, 12)) +
  scale_y_continuous(limits = c(6, 12)) +
  theme_light() +
  theme(#axis.title.x = element_markdown(family = "serif", size = 16),
        #axis.title.y = element_markdown(family = "serif", size = 16),
        axis.title = element_blank(),
        axis.text = element_text(family = "serif", size = 14))
#ggsave("figures/MLP_test_performance.jpg",dpi = 600, width = 8, height = 5)
```

### write key results to disk

```{r}
# write_rds(final_mlp_Tricho_nifH_abundance_wf, file = "data/key_model_results/final_mlp_Tricho_nifH_abundance_wf.RDS")
# 
# write_rds(mlp_Tricho_nifH_abundance_last_fit, file = "data/key_model_results/mlp_Tricho_nifH_abundance_last_fit.RDS")
```

## Comparing the performance of MLP and RF models on testing dataset

```{r}
collect_predictions(mlp_Tricho_nifH_abundance_last_fit) %>%
  mutate(model = "mlp") %>%
  bind_rows(collect_predictions(rf_Tricho_nifH_abundance_last_fit) %>% mutate(model = "rf")) #%>%
  #write_rds("data/model_test_performance.RDS")
```

```{r}
library(patchwork)
fig_test_performance <-
  fig_mlp_test_performance/fig_rf_test_performance 
fig_test_performance <-
  fig_test_performance + 
    plot_annotation(tag_levels = "a")& 
    theme(plot.tag = element_text(size = 20, face = "bold",family = "serif"),
          plot.tag.position = c(0.01, 1.02),
          )

#fig_test_performance <-
  wrap_elements(panel = fig_test_performance) +
    labs(tag = "Predicted *nifH* abundance (log10 copies m<sup>-2</sup>)") +
    theme(plot.tag = element_markdown(size = 16, family = "serif", angle = 90),
          plot.tag.position = c(0.25, 0.5))


# ggsave("figures/fig_test_performance.pdf",fig_test_performance, dpi = 600, units = "in", width = 7*1.4, height = 7*1.3)
```

> MLP model was selected for further analysis\
> 1. According to RMSE and R<sup>2</sup>, the performance of MLP and RF models were similar.\
> 2. As shown in `fig_test_performance`, RF model tends to overpredict nifH abudance at the low end, and underestimate at the high end.
