---
title: "Tricho_nifH_existence"
output: html_document
date: "2024-11-12"
---

> -   In this Rmarkdown file, random forest (RF) models built and optimized to predict the global *Trichodesmium* distribution (presence-absence) using `tidymodels`.
> -   It took \~ 15 min to execute this script on our server.

> `tidymodels` tutorials:\
> - [Julia Silge's blog](https://juliasilge.com/blog/sf-trees-random-tuning/)\
> - [Tidy Modeling with R](https://www.tmwr.org)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(tidymodels)
library(doParallel)
registerDoParallel() #  On our server, 80 cores were used
# getDoParWorkers() # check how many cores are being used for parallel processing
```

```{r}
load("data/data_R_exported/DS_prepared_for_model_building.RData") 

# See `Dataset_preparation.R` for information of `DS_prepared_for_model_building.RData`
```

## Training, testing and validation dataset

```{r data prep}
set.seed(4)
Tricho_nifH_existence_data_split <- 
  DS_for_ML_building_Tricho_nifH_existence %>%
  drop_na() %>%
  mutate(Trichodesmium_nifH_existence_regrid = if_else(Trichodesmium_nifH_existence_regrid, "yes", "no")) %>%
  mutate(Trichodesmium_nifH_existence_regrid = factor(Trichodesmium_nifH_existence_regrid)) %>%
  initial_split(strata = "Trichodesmium_nifH_existence_regrid")

Tricho_nifH_existence_data_train <-
  training(Tricho_nifH_existence_data_split)

Tricho_nifH_existence_data_test <-
  testing(Tricho_nifH_existence_data_split)

set.seed(234)
Tricho_nifH_existence_data_folds <- vfold_cv(Tricho_nifH_existence_data_train, v = 10, repeats = 5)

```

## Set up the `workflow` for model building, hyperparameter tuning and prediction.

```{r workflow set up}
rf_Tricho_nifH_existence_rec <-
  recipe(Trichodesmium_nifH_existence_regrid ~ ., data = Tricho_nifH_existence_data_train) %>%
  update_role(lat_regrid, lon_regrid, Sampling_month, new_role = "ID")

# Three hyperparameters were tuned: 
# 1. mtry: Randomly Selected Predictors
# 2. trees: number of Trees
# 3. min_n: Minimal Node Size
rf_Tricho_nifH_existence_spec <-
  rand_forest(
    mtry = tune(),
    trees = tune(),
    min_n = tune()
  ) %>%
  set_mode("classification") %>%
  set_engine("ranger")

rf_Tricho_nifH_existence_wf <-
  workflow() %>%
  add_recipe(rf_Tricho_nifH_existence_rec) %>%
  add_model(rf_Tricho_nifH_existence_spec)
```

## Tuning hyperparameters

### coarse tuning

```{r coarse tune in 40 seconds}
set.seed(345)
rf_Tricho_nifH_existence_tune <- tune_grid(
  rf_Tricho_nifH_existence_wf,
  resamples = Tricho_nifH_existence_data_folds,
  grid = 50
)
autoplot(rf_Tricho_nifH_existence_tune)
```

### fine tuning

```{r fine tune in ~ 90 seconds}
#based on the results of the last chunk,  hyperparameters
# were further tuned.
set.seed(12)
rf_Tricho_nifH_tune_grid <- 
  grid_space_filling(
    mtry(range = c(2, 14)),
    trees(range = c(50, 2000)),
    min_n(range = c(2, 20)),
    size = 100
  )

set.seed(14)
rf_Tricho_nifH_tune_fine <- tune_grid(
  rf_Tricho_nifH_existence_wf,
  resamples = Tricho_nifH_existence_data_folds,
  grid = rf_Tricho_nifH_tune_grid
)
autoplot(rf_Tricho_nifH_tune_fine)
```

```{r fine tune 1 in 60 s}
set.seed(124)
rf_Tricho_nifH_tune_grid1 <- 
  grid_space_filling(
    mtry(range = c(2, 10)),
    trees(range = c(50, 2000)),
    min_n(range = c(2, 10)),
    size = 100
  )

#a <- Sys.time()
set.seed(124)
rf_Tricho_nifH_tune_fine1 <- tune_grid(
  rf_Tricho_nifH_existence_wf,
  resamples = Tricho_nifH_existence_data_folds,
  grid = rf_Tricho_nifH_tune_grid1
)
#b <- Sys.time()
autoplot(rf_Tricho_nifH_tune_fine1)
```

```{r fine tune 2 in 180 s}
set.seed(1224)
rf_Tricho_nifH_tune_grid2 <- 
  grid_space_filling(
    mtry(range = c(2, 6)),
    trees(range = c(50, 2000)),
    min_n(range = c(2, 7)),
    size = 300
  )

a <- Sys.time()
set.seed(13424)
rf_Tricho_nifH_tune_fine2 <- tune_grid(
  rf_Tricho_nifH_existence_wf,
  resamples = Tricho_nifH_existence_data_folds,
  grid = rf_Tricho_nifH_tune_grid2
)
b <- Sys.time()
autoplot(rf_Tricho_nifH_tune_fine2)
```

## Selecting the best RF model

```{r select best rf model}
# select the best RF model
final_rf_Tricho_nifH_existence_wf <- 
  rf_Tricho_nifH_existence_wf %>%
  finalize_workflow(select_best(rf_Tricho_nifH_tune_fine2, metric = "roc_auc"))

# performance on testing dataset
set.seed(22)
rf_existence_last_fit <-
  last_fit(final_rf_Tricho_nifH_existence_wf, Tricho_nifH_existence_data_split)
collect_metrics(rf_existence_last_fit)
```

## Global prediciton of *Tricho* distribution

```{r Tricho nifH existence prediction}
set.seed(13424)
rf_Tricho_nifH_existence_prediction <-  
  predict(rf_existence_last_fit$.workflow[[1]], Predictors_for_model_prediction)
```

## write key results to disk

`final_rf_Tricho_nifH_existence_wf.RDS`, `rf_existence_last_fit.RDS` and `rf_Tricho_nifH_existence_prediction.RDS` were uploaded.

```{r}
# write_rds(final_rf_Tricho_nifH_existence_wf,
#   file = "data/key_model_results/final_rf_Tricho_nifH_existence_wf.RDS")
# write_rds(rf_existence_last_fit,
#   file = "data/key_model_results/rf_existence_last_fit.RDS")
# write_rds(rf_Tricho_nifH_existence_prediction,
#   file = "data/key_model_results/rf_Tricho_nifH_existence_prediction.RDS")
```
